Efficient and Scalable Bayesian Bipartite Matching with Fast Beta Linkage (fabl)

### 2021-12-20 methods_paper/fabl-paper.tex
updated BA submission

\title{Efficient and Scalable Bayesian Bipartite Matching with Fast Beta Linkage  (fabl)}
\author{Brian Kundinger\textsuperscript{a} \and
  Jerome Reiter\textsuperscript{a} \and 
  Rebecca C.~Steorts \textsuperscript{b}}
\date{
 \textsuperscript{a} Department of Statistical Science, Duke University \\
 \textsuperscript{b} Department of Statistical Science and Computer Science, Duke University\\Principal Mathematical Statistician, United States Census Bureau\\[2ex]
  \today}

@@

\bigskip
\begin{abstract}
Recently, researchers have developed Bayesian versions of the Fellegi Sunter model for record linkage. These have the crucial advantage of quantifying uncertainty from imperfect linkages.  However, current implementations of Bayesian Fellegi Sunter models are computationally intensive, making them challenging to use on larger-scale record linkage tasks.  We propose and investigate a variation on Bayesian Fellegi Sunter models that we call fast beta linkage, or \texttt{fabl}.  Specifically, in \texttt{fabl} we propose independent prior distributions over the matching space, allowing us to use hashing techniques that reduce computational overhead. This also allows us to complete pairwise record comparisons over large datasets through parallel computing and reduce memory costs through a new technique called storage efficient indexing. Through simulations and two case studies, we show that \texttt{fabl} has markedly increased speed with minimal loss of accuracy.

@@

where
%$w_{ij} = \left(\frac{\prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}}{\prod_{f=1}^{F}\prod_{l=1}^{L_f} u_{fl}}\right)^{I(\gamma_{ij}^f = l)I_{obs}(\gamma_{ij}^f)} = \frac{P(\boldsymbol{\gamma_{ij}}|Z_j = i)}{P(\boldsymbol{\gamma_{ij}} |Z_j \neq i)}$.
$$w_{ij} = \prod_{f=1}^{F}\prod_{l = 1}^{L_f} \left(\frac{m_{fl}}{u_{fl}}\right)^{I(\gamma_{ij}^f = l)I_{obs}(\gamma_{ij}^f)}$$.

@@

With the likelihood in this form, we can derive the full conditional
\begin{align*}
	p(Z_j|\Gamma_{.j}, \Phi, \pi) &\propto p(\Gamma_{.j}| Z_j, \Phi) P(Z_j|\pi) \\
	&\propto \left(\sum_{i=1}^{n_1}w_{ij}\bm{1}_{z_j = i} + \bm{1}_{z_j = n_1 + 1}\right)\left(\pi\sum_{i=1}^{n_1}\frac{1}{n_1}\bm{1}_{z_j = i} + (1-\pi)\bm{1}_{z_j = n_1 + 1}\right) \\	
	&= \frac{\pi}{n_1}\sum_{i=1}^{n_1}w_{ij}\bm{1}_{z_j = i} + (1-\pi)\bm{1}_{z_j = n_1 + 1}
\end{align*}
Thus we have
$$Z_j^{(s+1)} | \Phi, \Gamma, \pi \propto \begin{cases} 
	\frac{\pi}{n_1}w_{ij}   & z_j \leq n_1; \\
	1-\pi &  z_j  = n_1 + 1 \\
\end{cases}$$
Finally, we integrate over the posterior of \(\pi\) and rearrange terms to produce the final full conditional:

\[p\left(Z_j^{(s+1)}  = i| \Phi, \bm{Z^{(s)}}\right) \propto
\begin{cases} 
	w_{ij}  & i \leq n_1; \\
	n_1 \frac{n_2 - n_{12}(\bm{Z}) + \beta_{\pi}}{n_{12}(\bm{Z}) + \alpha_{\pi}} & i  = n_1 + 1 \\
\end{cases}\]

@@

\[L(\hat{Z_j}, Z_j)=\begin{cases} 
	0  & \text{if } Z_j = \hat{Z_j}; \\
	\theta_R,  & \text{if } \hat{Z_j} = R; \\
	\theta_{10},  & \text{if } Z_j \leq n_A,\hat{Z_j} = n_A + 1 ; \\
	\theta_{01},  & \text{if } Z_j = n_A + 1,\hat{Z_j} \leq n_A ; \\
	\theta_{11},  & \text{if } Z_j, \hat{Z}_j \leq n_A, Z_j \neq \hat{Z_j} ; \\
\end{cases}\] Here, \(\theta_R\) is the loss from not making a decision
on the linkage status, \(\theta_{10}\) is the loss from a false
non-match, \(\theta_{01}\) is the loss from a false match, and
\(\theta_{11}\) is the loss from the special case of a false match in
which the record has a true match other than the one estimated by the
model.

We seek to minimize the posterior expected loss, given by 
$$\mathbb{E}[L(\hat{\bm{Z}}, \bm{Z})|\Gamma^{obs}] = \sum_{\bm{Z}} L(\hat{Z_j}, Z_j)P(\bm{Z}|\Gamma^{obs}).$$
Here, we set $\theta_R = \infty, \theta_{10} = 1, \theta_{01} = 1, \theta_{11} = 2$, inducing the intuitive decision rule

@@

Since the Gibbs sampler does not strictly enforce one-to-one matching, it is possible for this Bayes estimate to link multiple records in $\bm{X_2}$ to one record in $\bm{X_1}$. It is often desirable to see such matches, because they provide evidence of model misspecification; specifically, large numbers of ``many to one'' matches indicate that the assumption of no duplicates within files may not be reasonable. We will explore this issue through the El Salvador homicide case study.

To achieve a Bayes estimate that fulfills one-to-one matching requirement, we simply minimize the expected loss subject to the constraint that $\hat{Z}_j \neq \hat{Z}_{j'}$ for all $j \neq j'$. In the event that we have two records $j$ and $j'$ such that both $P(Z_j = i |\Gamma) > \frac{1}{2}$ and $ P(Z_{j'} = i |\Gamma) > \frac{1}{2}$, we accept the match with the highest posterior probability, and declare the other to have no match. A similar approach appears in the most probable maximal matching sets used by \cite{steorts_bayesian_2016} to match records to latent entities.

@@

To increase the speed of the Gibbs sampler and expand the scale of the linkage tasks we can undertake, we propose hashing methods similar to those used by \cite{enamorado2019using} in the creation of \texttt{fastlink}, a fast and scalable implementation of the Fellegi Sunter method. The key insight is to recognize that record pairs contribute to posterior calculations only through the agreement pattern of the $\gamma_{ij}$ vector. To make this more precise, let $h_1, \ldots, h_P$ denote the unique agreement patterns, and collect these unique patterns in the set $\mathcal{P} = \{h_1, \ldots, h_P\}$. Here, 
$P = |\mathcal{P}|$ is the total number of unique agreement
patterns. \(P\) is bounded above by \(\prod_{f=1}^F L_f\) which does not depend on \(n_1\) or \(n_2\). In this context, the integers $\{1, \ldots, P\}$ serve as hashed values that encode the same information as the orignial vectors themselves. Whenever possible, we conduct calculations over these \(P\) agreement patterns, instead of the typical \(n_1 \times n_2\) record pairs. Additionally, we store ``one-hot encodings'' of these patterns rather than the original Fellegi Sunter comparison vectors, to aid in vectorized computations.

@@

With all record pairs converted to hashed values, we can create a more compact nested list 
$$\mathcal{R} = \left\{\{r_{j_p} \}_{p=1}^{P} \right\}_{j = 1}^{n_2}$$
where $r_{j_p} = \{i \in \bm{X_1} | (i, j) \in h_p\}$. This representation is useful because it allows us to easily compute sufficient statistics
$$H_{j_p} = \sum_{i} I((i, j) \in h_p) = ||r_{j_p}||$$
and
$$H_p =  \sum_{i, j} I((i, j) \in h_p) = \sum_{j} H_{j_p}$$
For convenience, denote this set of sufficient statistics $\mathcal{H} = \left\{\{H_{j_p}\}, \{H_{p}\}\right\}.$ As we will shown in Section \ref{efficient-posterior}, all posterior calculations are conducted with these sufficient statistics. Note that $\mathcal{P}$, $\mathcal{R}$, and $\mathcal{H}$  fully characterize the comparison matrix $\Gamma$ with no loss of information, so we proceed with $\tilde{\Gamma} = \{\mathcal{P}, \mathcal{R}, \mathcal{H}\}$ for posterior inference.

@@

The hashing procedure described above considerably reduces the memory
needed to store the comparison information. That is, instead of storing
\(n_1 \times n_2\) comparison vectors, we only store the
\(P\) unique vectors, and then \(n_1 \times n_2\) scalar quantities
relating record pairs to those vectors. However, even storing these
\(n_1 \times n_2\) scalar labels can become burdensome with large data.
Worse, the overwhelming majority of these labels relate to record pairs
that are clear non-matches.

To address this issue, we propose a new method called storage efficient indexing (SEI). In standard indexing, one decides a priori certain criteria that all true matching pairs must satisfy, 
and labels any record pairs that do not meet that criteria as
non-matches. For example, one might only consider pairs with a certain
similarity score on a field deemed to be important (like first name), or
only pairs that exactly match on a specified number of fields. While
generally chosen to be quite loose, establishing these criteria
requires knowledge of the problem and invites room for human error.

With SEI however, we can reduce the comparison space, and its associated storage costs, while avoiding these drawbacks. Observe that all record $i \in \bm{X}_1$ that share the same agreement pattern with $j \in \bm{x}_2$ have the same Fellegi Sunter weight $w_{ij}$, and therefore the same probability when sampling \(Z_j\). Thus, we know that records $i \in r_{j_p}$  such that $H_{j_p}$ is large are very unlikely to be sampled consistently enough to be deemed a match through the Bayes estimate. We know this regardless of the form of the agreement pattern itself, or its associated probabilities. Therefore, rather than store all of these unlikely record labels, we choose to store only a small number \(S\) of them in a new nested listed $\mathcal{R}^{\text{SEI}}$. Rather than storing \(n_1 \times n_2\) record labels, SEI allows us to store at most
\(n_2 \times P \times S\) labels, regardless of how large \(n_1\) might be. Posterior calculations still attribute the appropriate weight to all records through the summary statistics in $\mathcal{H}$, and thus we can proceed with posterior inference through the memory reduced $\tilde{\Gamma}^{\text{SEI}} = \{\mathcal{P}, \mathcal{R}^{\text{SEI}}, \mathcal{H}\}$. For simplicty, we suppress ``SEI'' in future notation, and use $\tilde{\Gamma}$ without loss of generality.

@@

Instead, we use one-hot encodings of the agreement patterns $\mathcal{P}$ for more efficient calculations. Denote
$H_p^m = \sum_{j=1}^{n_B} \bm{1}_{(Z_j, j) \in h_p}$ to be the
number of matching record pairs with agreement pattern $p$. We also can easily calculate the number of non-matching record pairs with agreement pattern $p$ as \(H_p^u = H_p - H_p^m\). Then, if $\bm{\alpha_0}$ and $\bm{\beta_0}$ are vectors of prior parameters for the $\bm{m}$ and $\bm{u}$ distributions respectively, the posterior update becomes

@@

$$P\left(h\left(Z_j^{(s+1)}\right) = p \mid \Phi, \bm{Z^{(s)}}\right) \propto
\begin{cases} 
	w_{p}\times H_{j_p}  & p \leq P; \\
	n_1 \frac{n_2 - n_{12}(\bm{Z}) + \beta_{\lambda}}{n_{12}(\bm{Z}) + \alpha_{\lambda}} &   p = P + 1 \\
\end{cases}$$
$$P\left(Z_j^{(s+1)} = i \mid h\left(Z_j^{(s+1)}\right) = p\right) = \begin{cases} 
	\frac{1}{H_{j_p}} & (i, j) \in h_p \\
	0 & \text{otherwise.} \\
\end{cases}$$

Lastly, we recognize that all posterior updates are governed by the agreement patterns of the record pairs rather than the record labels themselves. Thus we complete the entire Gibbs sampler first at the level of the \(P\) agreement patterns with the first equation above. After all iterations are complete, we can back-fill the records corresponding to the agreement patterns through the second equation.

We emphasize the computational gains of this split sampler: the first step is a sample from $P$ options, where $P$ does not scale with the size of the linkage task; and the second step is sampling uniformly at random, which is computationally simple even for large sets of candidate records. These changes drastically improve the speed of the sampler, and each can be parallelized if desired for additional computational gains. We provide a summary of the \texttt{fabl} method through pseudocode in Appendix \ref{app:fabl-psuedocode}. {\color{blue}We note here that this split sampler can be adapted for use with \texttt{BRL} as well, but would require updating $\{H_j\}$ to account for the previously matched records that no are longer candidate matches for the current record. \texttt{BRL} would see considerable speed up through the procedure, but this step would keep the complexity of the sampler at $O(n_A \times n_B)$.}

@@

\begin{proof}
To prove the computational complexity, we consider two steps --- constructing the comparison vectors and the Gibbs sampler. 
The computational complexity of all pairwise comparisons across the two databases $\bm{X}_1$ and $\bm{X}_2$ is $O(F n_1 n_2)$. The hashing procedure for all pairwise comparisons is also of complexity $O(F n_1 n_2)$. Since \texttt{fabl} is able to compute these comparisons in parallel, we can split these computations across $B$ equally sized partitions, so the complexity becomes $O(\frac{F}{B} n_1 n_2)$. There are then trivial computational costs associated with synthesizing summary statistics across these partitions.

@@

In our first simulation, we generate comparison vectors from pre-specified distributions so that we can easily increase the size of the linkage problem. The distributions used (see Table~\ref{Tab:distributions}) are meant to emulate the rates of agreement across first name, last name, day of birth, and month of birth.  For example, $u^{\text{month, agree}} = P(\text{Records have same birth-month} \mid \text{Nonmatch}) = \frac{1}{12}$. The parameters for year would vary largely by context, so we choose them here to adjust the desired difficulty of the linkage task. For simplicity, we consider only exact matching, so a vector (1, 0) corresponds to agreement and (0, 1) to disagreement. We simulate these data for different values of \(n_1\) and \(n_2\), and compare the run-time of \texttt{fabl} against \texttt{BRL}. Since we have 5 binary comparison fields, the number is unique patterns \(P\) is bounded above by \(2^5 = 32\), a bound which is consistently attained in the larger simulations.

@@

We note here that \texttt{BRL} is coded in C, which makes for unfair
comparison against \texttt{fabl}, currently built in R. However, the computational speed gains for \texttt{fabl} persist, especially for the larger simulations.  Additionally, although \texttt{fabl} is amenable to parallelization, this simulation was run on a single core. Implementing \texttt{fabl} in C++ with paralellization for the hashing step and sampling the matching status of the record pairs should lead to even more computational gains.

In Figure~\ref{fig:speed1}, we see that at low data size, \texttt{BRL} outperforms, but that \texttt{fabl} is significantly faster at handling larger data. In particular, run-time for \texttt{BRL} seems to grow quadratically (or linearly with the size of both $\bm{X}_1$ and $\bm{X}_2$) while run-time for \texttt{fabl} seems to grow linearly (in the size of $\bm{X}_2$).

\begin{figure}[t]
	\begin{center} \includegraphics[width=0.6\textwidth]{../notes/figures/sadinle_speed_plot2} 
		\caption{Run-time for BRL and fabl to run 1000 Gibbs iterations, including hashing step for fabl, for increasing values of both $n_A$ and $n_B$. We see near quadratic growth in runtime for BRL, and near linear growth for fabl.}\label{fig:speed1}
	\end{center}
\end{figure}

The above discussion suggests that for fixed \(n_2\), computation time
should remain mostly constant with growing \(n_1\). Simulation study
suggests that this is true. In Figure~\ref{fig:speed2} fixing \(n_2 = 500\), we
see linear growth for the run-time under \texttt{BRL} as \(n_1\)
increases, with much more static run-time under \texttt{fabl}. The
slight increases in run-time that we do see are due primarily to the
hashing step, which again can be run in parallel for large data.

\begin{figure}[h!]
	\begin{center} \includegraphics[width=0.6\textwidth]{../notes/figures/speed_plot_fixed_nB_slides} 
		\caption{Run-time for BRL and fabl to run 1000 Gibbs iterations, including hashing step for fabl, with increasing $n_A$, and $n_B$ fixed at 500. We see linear growth in runtime for BRL, and near constant runtime for fabl.}\label{fig:speed2}
	\end{center}
\end{figure}

@@

We replicate a simulation study from \cite{sadinle_bayesian_2017} to compare \texttt{fabl} against \texttt{BRL} on several synthetic datasets with varying amounts of error and overlap. To generate these data, clean records are first simulated from frequency tables for first name, last name, age, and occupation in Australia. Fields are then chosen for distortion uniformly at random. Names are subject to string insertions, deletions and substitutions, as well as common keyboard, phonetic, and optical recognition errors. Age and occupation are distorted through keyboard errors and missingness. For further details on the generation of these data, see \cite{christen_pudjijono2009} and \cite{christen_vatsalan2013}.

>>>>

\title{Efficient and Scalable Bayesian Bipartite Matching with Fast Beta Linkage  (fabl)}
\author{Brian Kundinger\textsuperscript{a} \and
  Jerome P. Reiter\textsuperscript{a} \and 
  Rebecca C.~Steorts \textsuperscript{b}}
\date{
 \textsuperscript{a} Department of Statistical Science, Duke University \\
 \textsuperscript{b} Department of Statistical Science and Computer Science, Duke University\\Principal Mathematical Statistician, United States Census Bureau\\[2ex]
  \today}

@@

\bigskip
\begin{abstract}
Recently, researchers have developed Bayesian versions of the Fellegi Sunter model for record linkage. These have the crucial advantage of quantifying uncertainty from imperfect linkages.  However, current implementations of Bayesian Fellegi Sunter models are computationally intensive, making them challenging to use on larger-scale record linkage tasks.  We propose a variation on Bayesian Fellegi Sunter models that we call fast beta linkage, or \texttt{fabl}.  Specifically, in \texttt{fabl} we use independent prior distributions over the matching space, allowing us to use hashing techniques that reduce computational overhead. This also allows us to complete pairwise record comparisons over large datasets through parallel computing and reduce memory costs through a new technique called storage efficient indexing. Through simulations and two case studies, we show that \texttt{fabl} has increased speed with minimal loss of accuracy.

@@

where
%$w_{ij} = \left(\frac{\prod_{f=1}^{F}\prod_{l=1}^{L_f} m_{fl}}{\prod_{f=1}^{F}\prod_{l=1}^{L_f} u_{fl}}\right)^{I(\gamma_{ij}^f = l)I_{obs}(\gamma_{ij}^f)} = \frac{P(\boldsymbol{\gamma_{ij}}|Z_j = i)}{P(\boldsymbol{\gamma_{ij}} |Z_j \neq i)}$.
$$w_{ij} = \prod_{f=1}^{F}\prod_{l = 1}^{L_f} \left(\frac{m_{fl}}{u_{fl}}\right)^{I(\gamma_{ij}^f = l)I_{obs}(\gamma_{ij}^f)}.$$

@@

With the likelihood in this form, we can derive the full conditional
\begin{align*}
	p(Z_j|\Gamma_{.j}, \Phi, \pi) &\propto p(\Gamma_{.j}| Z_j, \Phi) P(Z_j|\pi) \\
	&\propto \left(\sum_{i=1}^{n_1}w_{ij}\bm{1}_{z_j = i} + \bm{1}_{z_j = n_1 + 1}\right)\left(\pi\sum_{i=1}^{n_1}\frac{1}{n_1}\bm{1}_{z_j = i} + (1-\pi)\bm{1}_{z_j = n_1 + 1}\right) \\	
	&= \frac{\pi}{n_1}\sum_{i=1}^{n_1}w_{ij}\bm{1}_{z_j = i} + (1-\pi)\bm{1}_{z_j = n_1 + 1}.
\end{align*}
Thus, we have
$$Z_j^{(s+1)} | \Phi, \Gamma, \pi \propto \begin{cases} 
	\frac{\pi}{n_1}w_{ij}   & z_j \leq n_1; \\
	1-\pi &  z_j  = n_1 + 1. \\
\end{cases}$$
We integrate over \(\pi\) and rearrange terms to produce the final full conditional:

\[p\left(Z_j^{(s+1)}  = i| \Phi, \bm{Z^{(s)}}\right) \propto
\begin{cases} 
	w_{ij}  & i \leq n_1; \\
	n_1 \frac{n_2 - n_{12}(\bm{Z}) + \beta_{\pi}}{n_{12}(\bm{Z}) + \alpha_{\pi}} & i  = n_1 + 1. \\
\end{cases}\]

@@

\[L(\hat{Z_j}, Z_j)=\begin{cases} 
	0  & \text{if } Z_j = \hat{Z_j}; \\
	\theta_R,  & \text{if } \hat{Z_j} = R; \\
	\theta_{10},  & \text{if } Z_j \leq n_A,\hat{Z_j} = n_A + 1 ; \\
	\theta_{01},  & \text{if } Z_j = n_A + 1,\hat{Z_j} \leq n_A ; \\
	\theta_{11},  & \text{if } Z_j, \hat{Z}_j \leq n_A, Z_j \neq \hat{Z_j}. \\
\end{cases}\] Here, \(\theta_R\) is the loss from not making a decision
on the linkage status, \(\theta_{10}\) is the loss from a false
non-match, \(\theta_{01}\) is the loss from a false match, and
\(\theta_{11}\) is the loss from the special case of a false match in
which the record has a true match other than the one estimated by the
model.

%In this article, we set false matches and false non-matches to be equally costly, and thus set $\theta_{10} = \theta_{01} = 1$ and $\theta_{11} = 2$; when we want a full estimate of the linkage parameter, we set $\theta_R = \infty$, and when we want to leave some records unmatched by the algorithm, we set $\theta_R = 0.1$.

We seek to minimize the posterior expected loss, given by 
$$\mathbb{E}[L(\hat{\bm{Z}}, \bm{Z})|\Gamma^{obs}] = \sum_{\bm{Z}} L(\hat{Z_j}, Z_j)P(\bm{Z}|\Gamma^{obs}).$$

Here, we set $(\theta_{10}, \theta_{01}, \theta_{11}, \theta_R) = (1, 1, 2, \infty)$ inducing the intuitive decision rule

%$\theta_R = \infty, \theta_{10} = 1, \theta_{01} = 1, \theta_{11} = 2$ ,

@@

Since the Gibbs sampler does not strictly enforce one-to-one matching, it is possible for this Bayes estimate to link multiple records in $\bm{X_2}$ to one record in $\bm{X_1}$. To achieve a Bayes estimate that fulfills one-to-one matching requirement, we simply minimize the expected loss subject to the constraint that $\hat{Z}_j \neq \hat{Z}_{j'}$ for all $j \neq j'$. In the event that we have two records $j$ and $j'$ such that both $P(Z_j = i |\Gamma) > \frac{1}{2}$ and $ P(Z_{j'} = i |\Gamma) > \frac{1}{2}$, we accept the match with the highest posterior probability, and declare the other to have no match. A similar approach appears in the most probable maximal matching sets used by \cite{steorts_bayesian_2016} to match records to latent entities.

%It is often desirable to see such matches, because they provide evidence of model misspecification; specifically, large numbers of ``many to one'' matches indicate that the assumption of no duplicates within files may not be reasonable. We will explore this issue through the El Salvador homicide case study.

@@

To increase the speed of the Gibbs sampler and expand the scale of the linkage tasks we can undertake, we propose hashing methods similar to those used by \cite{enamorado2019using} in the creation of \texttt{fastlink}, a fast and scalable implementation of the Fellegi Sunter method. The key insight is to recognize that record pairs contribute to posterior calculations only through the agreement pattern of the $\gamma_{ij}$ vector. To make this more precise, let $h_1, \ldots, h_P$ denote the unique agreement patterns, and collect these unique patterns in the set $\mathcal{P} = \{h_1, \ldots, h_P\}$. Here, 
$P = |\mathcal{P}|$ is the total number of unique agreement
patterns. \(P\) is bounded above by \(\prod_{f=1}^F L_f\) which does not depend on \(n_1\) or \(n_2\). In this context, the integers $\{1, \ldots, P\}$ serve as hashed values that encode the same information as the orignial vectors themselves. Whenever possible, we conduct calculations over these \(P\) agreement patterns, instead of the typical \(n_1 \times n_2\) record pairs. Additionally, we store ``one-hot encodings'' of these patterns rather than the original Fellegi Sunter comparison vectors, to aid in vectorized computations. See Appendix ~\ref{app:ohe} for details.

@@

With all record pairs converted to hashed values, we can create a more compact nested list 
$$\mathcal{R} = \left\{\{r_{j_p} \}_{p=1}^{P} \right\}_{j = 1}^{n_2}$$
where $r_{j_p} = \{i \in \bm{X_1} | (i, j) \in h_p\}$. This representation is useful because it allows us to easily compute sufficient statistics
$$H_{j_p} = \sum_{i} I((i, j) \in h_p) = ||r_{j_p}||$$
and
$$H_p =  \sum_{i, j} I((i, j) \in h_p) = \sum_{j} H_{j_p}.$$
For convenience, denote this set of sufficient statistics $\mathcal{H} = \left\{\{H_{j_p}\}, \{H_{p}\}\right\}.$ As we will show in Section \ref{efficient-posterior}, all posterior calculations are conducted with these sufficient statistics. Note that $\mathcal{P}$, $\mathcal{R}$, and $\mathcal{H}$  fully characterize the comparison matrix $\Gamma$ with no loss of information, so we proceed with $\tilde{\Gamma} = \{\mathcal{P}, \mathcal{R}, \mathcal{H}\}$ for posterior inference.

@@

The hashing procedure described above considerably reduces the memory needed to store the comparison information. That is, instead of storing \(n_1 \times n_2\) comparison vectors, we only store the \(P\) unique vectors, and then \(n_1 \times n_2\) scalar quantities relating record pairs to those vectors. However, even storing these \(n_1 \times n_2\) scalar labels can become burdensome with large databases. Additionally, the overwhelming majority of these labels relate to record pairs that are clear non-matches.

To address this issue, we propose storage efficient indexing (SEI). In standard indexing, one decides a priori certain criteria that all true matching pairs must satisfy, and labels any record pairs that do not meet that criteria as non-matches. For example, one might only consider pairs with a certain similarity score on a field deemed to be important (like first name), or only pairs that exactly match on a specified number of fields. While generally chosen to be quite loose, establishing these criteria requires knowledge of the problem and invites room for human error.

With SEI, however, we can reduce the comparison space, and its associated storage costs, while avoiding these drawbacks. Observe that all records $i \in \bm{X}_1$ that share the same agreement pattern with $j \in \bm{x}_2$ have the same Fellegi Sunter weight $w_{ij}$, and therefore the same probability when sampling \(Z_j\). Thus, we know that records $i \in r_{j_p}$  such that $H_{j_p}$ is large are very unlikely to be sampled consistently enough to be deemed a match through the Bayes estimate. We know this regardless of the form of the agreement pattern itself, or its associated probabilities. Therefore, rather than store all of these unlikely record labels, we choose to store only a small number \(S\) of them in a new nested listed $\mathcal{R}^{\text{SEI}}$. Rather than storing \(n_1 \times n_2\) record labels, SEI allows us to store at most
\(n_2 \times P \times S\) labels, regardless of how large \(n_1\) might be. Posterior calculations still attribute the appropriate weight to all records through the summary statistics in $\mathcal{H}$, and thus we can proceed with posterior inference through the memory reduced $\tilde{\Gamma}^{\text{SEI}} = \{\mathcal{P}, \mathcal{R}^{\text{SEI}}, \mathcal{H}\}$. For simplicty, we suppress ``SEI'' in future notation, and use $\tilde{\Gamma}$ without loss of generality.

@@

Instead, we use one-hot encodings of the agreement patterns $\mathcal{P}$ for more efficient calculations. Denote $H_p^m = \sum_{j=1}^{n_B} \bm{1}_{(Z_j, j) \in h_p}$ to be the number of matching record pairs with agreement pattern $p$. The number of non-matching record pairs with agreement pattern $p$ is \(H_p^u = H_p - H_p^m\). Then, if $\bm{\alpha_0}$ and $\bm{\beta_0}$ are vectors of prior parameters for the $\bm{m}$ and $\bm{u}$ distributions respectively, the posterior update becomes

@@

%$$P\left(h\left(Z_j^{(s+1)}\right) = p \mid \Phi, \bm{Z^{(s)}}\right) \propto
%\begin{cases} 
%	w_{p}\times H_{j_p}  & p \leq P; \\
%	n_1 \frac{n_2 - n_{12}(\bm{Z}) + \beta_{\lambda}}{n_{12}(\bm{Z}) + \alpha_{\lambda}} &   p = P + 1 \\
%\end{cases}$$
%$$P\left(Z_j^{(s+1)} = i \mid h\left(Z_j^{(s+1)}\right) = p\right) = \begin{cases} 
%	\frac{1}{H_{j_p}} & (i, j) \in h_p \\
%	0 & \text{otherwise.} \\
%\end{cases}$$
%
\begin{align}
	\label{eqn:gibbs1}
	P\left(h\left(Z_j^{(s+1)}\right) = p \mid \Phi, \bm{Z^{(s)}}\right) \propto
	\begin{cases} 
		w_{p}\times H_{j_p}  & p \leq P; \\
		n_1 \frac{n_2 - n_{12}(\bm{Z}) + \beta_{\lambda}}{n_{12}(\bm{Z}) + \alpha_{\lambda}} &   p = P + 1 \\
	\end{cases}
\end{align}
\begin{align}
	\label{eqn:gibbs2}
	P\left(Z_j^{(s+1)} = i \mid h\left(Z_j^{(s+1)}\right) = p\right) = \begin{cases} 
	\frac{1}{H_{j_p}} & (i, j) \in h_p \\
	0 & \text{otherwise.} \\
\end{cases}
\end{align}

Lastly, we recognize that all posterior updates are governed by the agreement patterns of the record pairs rather than the record labels themselves. Thus we complete the entire Gibbs sampler first at the level of the \(P\) agreement patterns with (\ref{eqn:gibbs1}) above. After all iterations are complete, we can back-fill the records corresponding to the agreement patterns through (\ref{eqn:gibbs2}).

We emphasize the computational gains of this split sampler: the first step is a sample from $P$ options, where $P$ does not scale with the size of the linkage task; and the second step is sampling uniformly at random, which is computationally simple even for large sets of candidate records. These changes drastically improve the speed of the sampler, and each can be parallelized if desired for additional computational gains. We provide a summary of the \texttt{fabl} method through pseudocode in Appendix \ref{app:fabl-psuedocode}. We note here that this split sampler can be adapted for use with \texttt{BRL} as well, but would require updating $\{H_j\}$ to account for the previously matched records that no are longer candidate matches for the current record. \texttt{BRL} would see considerable speed up through the procedure, but computation time would still scale with the size of the larger dataset.

@@

\begin{proof}
To prove the computational complexity, we consider two steps --- constructing the comparison vectors and the Gibbs sampler. 
The computational complexity of all pairwise comparisons across the two databases $\bm{X}_1$ and $\bm{X}_2$ is $O(F n_1 n_2)$. The hashing procedure for all pairwise comparisons is also of complexity $O(F n_1 n_2)$. With $B$ processors available, we can split these computations across $B$ equally sized partitions and compute these comparisons in parallel, so the complexity becomes $O(\frac{F}{B} n_1 n_2)$. There are then trivial computational costs associated with synthesizing summary statistics across these partitions.

@@

In our first simulation, we generate comparison vectors from pre-specified distributions so that we can easily increase the size of the linkage problem. The distributions used (see Table~\ref{Tab:distributions}) are meant to emulate the rates of agreement across first name, last name, day of birth, and month of birth.  For example, $u^{\text{month, agree}} = P(\text{Records have same birth-month} \mid \text{Nonmatch}) = \frac{1}{12}$. The parameters for year would vary largely by context, so we choose them here to adjust the desired difficulty of the linkage task. For simplicity, we consider only exact matching, so a vector (1, 0) corresponds to agreement and (0, 1) to disagreement. We simulate these data for different values of \(n_1\) and \(n_2\), and compare the run-time of \texttt{fabl} against \texttt{BRL}. Since we have 5 binary comparison fields, the number is unique patterns \(P\) is bounded above by \(2^5 = 32\), a bound which is consistently attained in simulations with more records.

@@

We note here that the implementation of \texttt{BRL} that we use is coded in C \citep{sadinle_bayesian_2017}.  In contrast, we use non-optimized code written in R for \texttt{fabl}.  While this complicates comparisons, and indeed disfavors fabl, the computational speed gains for \texttt{fabl} are still evident, especially for larger sample sizes.  Additionally, although \texttt{fabl} is amenable to parallelization, this simulation was run on a single core. Implementing \texttt{fabl} in C++ with paralellization for the hashing step and sampling the matching status of the record pairs should lead to even more computational gains.

In Figure~\ref{fig:speed1}, \texttt{BRL} is faster than \texttt{fabl} for low sample sizes, but that \texttt{fabl} is significantly faster at handling larger data. In particular, run-time for \texttt{BRL} seems to grow quadratically (or linearly with the size of both $\bm{X}_1$ and $\bm{X}_2$) while run-time for \texttt{fabl} seems to grow linearly (in the size of $\bm{X}_2$).

\begin{figure}[t]
	\begin{center} \includegraphics[width=0.6\textwidth]{../notes/figures/sadinle_speed_plot2} 
		\caption{Run-time for BRL and fabl to run 1000 Gibbs iterations, including hashing step for fabl, for increasing values of both $n_A$ and $n_B$. We see near quadratic growth in run-time for BRL, and near linear growth for fabl.}\label{fig:speed1}
	\end{center}
\end{figure}

The above discussion suggests that for fixed \(n_2\), computation time for \texttt{fabl} should remain mostly constant with growing \(n_1\). In the simulation results displayed in Figure~\ref{fig:speed2} fixing \(n_2 = 500\), we see linear growth for the run-time under \texttt{BRL} as \(n_1\) increases, with much more static run-time under \texttt{fabl}. The slight increases in run-time for \texttt{fabl} are due primarily to the hashing step, which again can be run in parallel for large data.

\begin{figure}[h!]
	\begin{center} \includegraphics[width=0.6\textwidth]{../notes/figures/speed_plot_fixed_nB_slides} 
		\caption{Run-time for BRL and fabl to run 1000 Gibbs iterations, including hashing step for fabl, with increasing $n_A$, and $n_B$ fixed at 500. We see linear growth in run-time for BRL, and near constant run-time for fabl.}\label{fig:speed2}
	\end{center}
\end{figure}

@@

We replicate a simulation study from \cite{sadinle_bayesian_2017} to compare \texttt{fabl} against \texttt{BRL} on several synthetic datasets with varying amounts of error and overlap (the number of records in common across files). Following methods proposed by \cite{christen_pudjijono2009} and \cite{christen_vatsalan2013}, clean records are first simulated from frequency tables for first name, last name, age, and occupation in Australia. Fields are then chosen for distortion uniformly at random. Names are subject to string insertions, deletions and substitutions, as well as common keyboard, phonetic, and optical recognition errors. Age and occupation are distorted through keyboard errors and missingness.

We create comparison vectors according to the default settings of the \texttt{compareRecords} function from the \texttt{BRL} package, shown in Table~\ref{Tab:sadinle_simulation_cutoffs}. Each simulation identifies matched individuals between two datasets, each with 500 records. We conduct linkage when matching records exhibit 1, 2, and 3 errors across the four fields, and when there are 50, 250, and 450 individuals in common across datasets. We use flat priors for all $m$ and $u$ parameters, run the Gibbs Sampler for 1000 iterations, and discard the first 100 as burn-in. We calculate Bayes estimates of the linkage structure using the losses $\theta_R = \infty, \theta_{10} = 1, \theta_{01} = 1, \theta_{11} = 2$. Traceplots for parameters of interest for one example simulation are provided in Appendix \ref{app:appendix-sim}; they show no obvious concern over MCMC convergence.
